---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Continuous beta-binomial distribution

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/cbbinom)](https://CRAN.R-project.org/package=cbbinom)
[![R-CMD-check](https://github.com/zhuxr11/cbbinom/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/zhuxr11/cbbinom/actions/workflows/R-CMD-check.yaml)
[![Download stats](https://cranlogs.r-pkg.org/badges/grand-total/cbbinom)](https://CRAN.R-project.org/package=cbbinom)
<!-- badges: end -->

**Package**: [*cbbinom*](https://github.com/zhuxr11/cbbinom) `r pkgload::pkg_version()`<br />
**Author**: Xiurui Zhu<br />
**Modified**: `r file.info("README.Rmd")$mtime`<br />
**Compiled**: `r Sys.time()`

The goal of `cbbinom` is to implement continuous beta-binomial distribution.

## Installation

You can install the released version of `cbbinom` from [CRAN](https://cran.r-project.org/) with:

``` r
install.packages("cbbinom")
```

Alternatively, you can install the developmental version of `cbbinom` from [github](https://github.com/) with:

``` r
remotes::install_github("zhuxr11/cbbinom")
```

## Introduction to continuous beta-binomial distribution

The continuous beta-binomial distribution spreads the standard probability mass of beta-binomial distribution at `x` to an interval `[x, x + 1]` in a continuous manner. This can be validated via the following plot, where we can see that the cumulative distribution function (CDF) of the continuous beta-binomial distribution at `x + 1` equals to that of the beta-binomial distribution at `x`.

```{r library, eval=FALSE}
library(cbbinom)
```

```{r, include=FALSE}
pkgload::load_all()
```

```{r cbbinom-vs-bbinom-p, fig.width=7, fig.height=5, fig.dpi=192}
# The continuous beta-binomial CDF, shift by -1
cbbinom_plot_x <- seq(-1, 10, 0.01)
cbbinom_plot_y <- pcbbinom(
  q = cbbinom_plot_x,
  size = 10,
  alpha = 2,
  beta = 4,
  ncp = -1
)
# The beta-binomial CDF
bbinom_plot_x <- seq(0L, 10L, 1L)
bbinom_plot_y <- extraDistr::pbbinom(
  q = bbinom_plot_x,
  size = 10L,
  alpha = 2,
  beta = 4
)
ggplot2::ggplot(mapping = ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_bar(
    data = data.frame(
      x = bbinom_plot_x,
      y = bbinom_plot_y
    ),
    stat = "identity"
  ) +
  ggplot2::geom_point(
    data = data.frame(
      x = cbbinom_plot_x,
      y = cbbinom_plot_y
    )
  ) +
  ggplot2::scale_x_continuous(
    n.breaks = diff(range(cbbinom_plot_x))
  ) +
  ggplot2::theme_bw() +
  ggplot2::labs(y = "CDF(x)")
```

However, the central density at `x + 1/2` of the continuous beta-binomial distribution may not equal to the corresponding probability mass at `x`, especially around the summit and to the right (since `alpha < beta`).

```{r cbbinom-vs-bbinom-d}
# The continuous beta-binomial CDF, shift by -1/2
cbbinom_plot_x_d <- seq(-1/2, 10 + 1/2, 0.01)
cbbinom_plot_y_d <- dcbbinom(
  x = cbbinom_plot_x_d,
  size = 10,
  alpha = 2,
  beta = 4,
  ncp = -1/2
)
# The beta-binomial CDF
bbinom_plot_x <- seq(0L, 10L, 1L)
bbinom_plot_y_d <- extraDistr::dbbinom(
  x = bbinom_plot_x,
  size = 10L,
  alpha = 2,
  beta = 4
)
ggplot2::ggplot(mapping = ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_bar(
    data = data.frame(
      x = bbinom_plot_x,
      y = bbinom_plot_y_d
    ),
    stat = "identity"
  ) +
  ggplot2::geom_point(
    data = data.frame(
      x = cbbinom_plot_x_d,
      y = cbbinom_plot_y_d
    )
  ) +
  ggplot2::scale_x_continuous(
    n.breaks = diff(range(bbinom_plot_x))
  ) +
  ggplot2::theme_bw() +
  ggplot2::labs(y = "CDF(x)")
```

For larger sizes, you may need higher precision than double for accuracy, at the cost of computational speed. When computing probability density, it is advisable to use a higher accuracy level than the one sufficient to compute the corresponding cumulative probability (e.g. 20 -> 25), since numerical derivative taken to compute the probability density may be more sensitive to computational errors.

```{r cbbinom-prec-pd, fig.width=7, fig.height=5, fig.dpi=192}
cbbinom_plot_prec_x <- seq(0, 41, 0.1)
# Compute CDF at default (double) precision level
system.time(pcbbinom_plot_prec0_y <- pcbbinom(
  q = cbbinom_plot_prec_x,
  size = 40,
  alpha = 2,
  beta = 4,
  ncp = -1,
  prec = NULL
))
ggplot2::ggplot(data = data.frame(x = cbbinom_plot_prec_x,
                                  y = pcbbinom_plot_prec0_y),
                mapping = ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point() +
  ggplot2::theme_bw() +
  ggplot2::labs(y = "CDF(x)")
# Compute CDF at precision level 20
system.time(pcbbinom_plot_prec20_y <- pcbbinom(
  q = cbbinom_plot_prec_x,
  size = 40,
  alpha = 2,
  beta = 4,
  ncp = -1,
  prec = 20
))
ggplot2::ggplot(data = data.frame(x = cbbinom_plot_prec_x,
                                  y = pcbbinom_plot_prec20_y),
                mapping = ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point() +
  ggplot2::theme_bw() +
  ggplot2::labs(y = "CDF(x)")
# Compute PDF at precision level 20
system.time(dcbbinom_plot_prec20_y <- dcbbinom(
  x = cbbinom_plot_prec_x,
  size = 40,
  alpha = 2,
  beta = 4,
  ncp = -1,
  prec = 20
))
ggplot2::ggplot(data = data.frame(x = cbbinom_plot_prec_x,
                                  y = dcbbinom_plot_prec20_y),
                mapping = ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point() +
  ggplot2::theme_bw() +
  ggplot2::labs(y = "PDF(x)")
# Compute PDF at precision level 25
system.time(dcbbinom_plot_prec25_y <- dcbbinom(
  x = cbbinom_plot_prec_x,
  size = 40,
  alpha = 2,
  beta = 4,
  ncp = -1,
  prec = 25
))
ggplot2::ggplot(data = data.frame(x = cbbinom_plot_prec_x,
                                  y = dcbbinom_plot_prec25_y),
                mapping = ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point() +
  ggplot2::theme_bw() +
  ggplot2::labs(y = "PDF(x)")
```


## Examples of continuous beta-binomial distribution

As the probability distributions in `stats` package, `cbbinom` provides a full set of density, distribution function, quantile function and random generation for the continuous beta-binomial distribution.

```{r cbbinom-funs}
# Density function
dcbbinom(x = 5, size = 10, alpha = 2, beta = 4)
# Distribution function
(test_val <- pcbbinom(q = 5, size = 10, alpha = 2, beta = 4))
# Quantile function
qcbbinom(p = test_val, size = 10, alpha = 2, beta = 4)
# Random generation
set.seed(1111L)
rcbbinom(n = 10L, size = 10, alpha = 2, beta = 4)
```

For mathematical details, please check the details section of `?cbbinom`.

## `Rcpp` implementation of `stats::uniroot()`

As a bonus, `cbbinom` also exports an `Rcpp` implementation of `stats::uniroot()` function, which may come in handy to solve equations, especially the monotonic ones used in quantile functions. Here is an example to calculate `qnorm` from `pnorm` in `Rcpp`.

```{Rcpp uniroot, eval=FALSE}
#include <iostream>
#include "cbbinom.h"
using namespace cbbinom;

// Define a functor as pnorm() - p
class PnormEqn: public UnirootEqn
{
private:
  double mu;
  double sd;
  double p;
public:
  PnormEqn(const double mu_, const double sd_, const double p_):
    mu(mu_), sd(sd_), p(p_) {}
  double operator () (const double& x) const override {
    return R::pnorm(x, this->mu, this->sd, true, false) - this->p;
  }
};

// Compute quantiles
int main() {
  double p = 0.975;  // Quantile
  PnormEqn eqn_obj(0.0, 1.0, 0.975);
  double tol = 1e-6;
  int max_iter = 10000;
  double q = cbbinom::cpp_uniroot(-1000.0, 1000.0, -p, 1.0 - p, &eqn_obj, &tol, &max_iter);
  std::cout << "Quantile at " << p << "is: " << q << std::endl;
  return 0;
}
```

